A spiking neural integrator model of the adaptive control of action by the anterior cingulate cortex

### 2013-06-10 paper/jneurosci2013.tex
Fourth version of the paper, with figures.
- This is the version submitted to JNeuroscience

\documentclass[11pt]{article}
% Change "article" to "report" to get rid of page number on title page
\usepackage{amsmath,amsfonts,amsthm,amssymb}
\usepackage{setspace}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{extramarks}
\usepackage{chngpage}
\usepackage{soul}
\usepackage[round]{natbib}
\usepackage[usenames,dvipsnames]{color}
\usepackage{graphicx,float,wrapfig}
\usepackage{listings}
\usepackage{framed}
\usepackage{inconsolata}
\usepackage{hyperref}

@@

\pagestyle{empty}

@@

\setlength\fboxrule{1pt}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Make title
\title{Adaptive control of action timing with a spiking neural integrator
  model of anterior cingulate cortex}
\date{}
\author{Trevor Bekolay, Chris Eliasmith, Mark Laubach}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

@@

\begin{abstract}
  Animals can learn to perform a simple reaction time task
  with a predictable foreperiod
  by predicting the time of the cue,
  achieving reaction times faster
  than through cue-responding alone.
  The outcome of the previous trial
  is one factor in determining
  whether an animal will attempt to time
  the foreperiod.
  The anterior cingulate cortex (ACC) has been implicated
  in enabling control of action timing
  depending on the outcome of the previous trial.
  Analysis of ACC activity suggests that
  neural integration is a key mechanism
  for adaptive control in precisely timed tasks.
  We show that a spiking neural circuit
  consisting of coupled neural integrators
  captures the neural dynamics
  of the experimentally recorded ACC.
  We use this coupled integrator model
  to adaptively control
  a spiking neural circuit that follows
  a cue-response strategy,
  and show that it performs the simple reaction time task
  with similar reaction times as experimental subjects.
  We then alter the model
  in a way consistent with
  how the aged brain degrades,
  and predict that aged subjects
  will adopt a cue-response strategy.
\end{abstract}

@@

The anterior cingulate cortex has been implicated
in a wide variety of functions,
including conflict monitoring
\citep{Aarts2009, Botvinick2004, VanVeen2004},
error monitoring \citep{Brown2005, Carter1998, Fecteau2003,
  Modirrousta2008, Rushworth2004},
anticipatory control \citep{Aarts2008, Koyama2001},
arousal of the sympathetic nervous system
\citep{Critchley2003, Luu2003},
and control of action timing \citep{Muir1996, Mulert2003,
  Naito2013, Narayanan2009, Risterucci2003}.
Many have proposed integrative or general models
that would allow the ACC to be involved
in several of these functions,
depending on the behavioral context
\citep{Botvinick2004, Luu2003}.
However, while there have been
attempts at modeling the ACC at the level
of artificial neural networks
\citep{Botvinick2004, Brown2005},
there are currently no mathematical
accounts of ACC function that would
explain the myriad of behaviors
modulated by the ACC.
Similarly, no existing models
have been implemented with spiking neurons,
and are therefore not directly comparable
to neural activity.

In this paper, we propose that a
two-dimensional dynamical system
can capture the activity of the ACC.
We test the model in a simple reaction time task
with a predictable foreperiod and
show that the mathematical model
enables precise timing of actions,
and changes behavior depending
on the outcome of the last trial.
We implement the dynamical system
in a spiking neural network
based on \citet{Singh2006}
and show that simulated neural activity
closely matches recorded ACC activity.
We show that the model
can be combined with
a motor control model,
resulting in behavioral changes
analogous to previous behavioral theories
of ACC function.
Finally, we damage the model in a way analogous
to aging, and predict that
aged subjects will adopt a more conservative
strategy due to a reduction of the
ability of the ACC to control action timing.

Our mathematical account of ACC function
is applied directly to
error monitoring and
the control of action timing.
However, we assert that the dynamical systems
approach is general and can be used
to explain other theories
of high-level ACC function.

\section{Simple reaction time task} \label{sec:simple-rt}

The simple reaction time task
serves as a minimal task in which
to examine the control of precisely timed actions.
In this task, the subject presses and holds down a lever
for a fixed amount of time (the ``foreperiod'';
1 second for this study).
After the foreperiod, a cue is presented
(e.g., an auditory tone).
The subject then has 0.6 seconds (the response window)
in which to release the lever.
If the lever is released in that time,
the trial is a success, and the subject is rewarded.
If the lever is released during the foreperiod,
the trial is classified as a ``premature error,''
and is penalized by a time-out.
If the lever is not released within the response window,
the trail is classified as a ``late error,''
and is also penalized by a time-out.
Figure~\ref{fig1} presents a schematic view of the task.

\section{Analysis of experimental data}

\citet{Narayanan2009} recorded
the anterior cingulate cortex (ACC) and motor cortex
of rodents performing this simple reaction-time task.
They used principal component analysis
on normalized spike densities
in order to determine
what is being represented in these brain areas
during this task.
As shown in Figure~\ref{fig2},
the first two principal components
found from analyzing the ACC data are large,
and are hypothesized to track
1) if the subject was performing the task, and
2) the relative amount of time the subject has been
performing the task.
\citeauthor{Narayanan2009} also found that the neural activity
in the ACC changed significantly depending on
the outcome of the previous trial;
the first two principal components
change significantly after errors,
and appear to contain information that an error occurred.
This information would be necessary
in order to adapt behavior based on the last trial.

Interestingly, in the post-correct case,
the first principal component
is highly correlated with the integral
of the second principal component,
and vice versa.
In the post-error case,
these correlations are significantly weaker.
This result suggests that
the ACC is performing some kind of integration..

\section{Double integrator model}

One model of neural integration
was proposed in the context of working memory
by \citet{Singh2006}.
Their model, called the \textit{double integrator} model,
uses multidimensional neurons
in integrative populations
to exhibit time-varying activity
during a vibrotactile discrimination task.
It relies on multiple coupled integrators,
whose activity resembles
the two strong principal components
in the ACC data.
The model employs
the Neural Engineering Framework (NEF; \citealp{Eliasmith2003})
to implement a spiking neural network
that captures the dynamics of the
integrative dynamical system
(see section \ref{sec:nef}).
We hypothesize that the \citeauthor{Singh2006} model
will exhibit the same neural dynamics
as the experimental ACC in the simple RT task.

\subsection{Dynamics}

The dynamics of the double integrator model
are captured in the following dynamics equations.
\begin{align*}
  \dot{x}_1 &= u \\
  \dot{x}_2 &= \alpha x_1,
\end{align*}
where $u$ represents an input signal
and $\alpha$ is the strength of
the connection between the two integrators.
This results in a two-dimensional system
in which $x_1$ integrates its input,
maintaining the accumulated value over time.
$x_2$ integrates the value of $x_1$.
\citet{Singh2006} have proposed that this system
enables time tracking;
because $x_2$ integrates $x_1$,
the amount of time elapsed since
$x_1$ changes can be determined
based on how much $x_2$ has changed.

In the simple reaction time task,
the input, $u$, is the activation of neurons
that effect the lever press.
The system is also modified by signals
representing the outcome of a particular trial;
on a correct trial, reward delivery ($R > 0$)
resets the system to a starting state.
On an error trail, a second input signal ($E < 0$)
drives $x_1$ to a low state.
When implemented in a spiking network,
this causes $x_2$ to saturate at a low value,
effectively breaking the integration,
as is seen in Figure~\ref{fig3} in post-error trails.
This results in the following dynamics.
\begin{align} \label{dint}
  \dot{x}_1 &= u + E - R x_1 \nonumber \\
  \dot{x}_2 &= \alpha x_1 - R x_2.
\end{align}
See Figure~\ref{fig4} for a graphical
depiction of the double integrator model.

In addition to the depicted system,
we inject a slowly oscillating signal into $x_1$.
This is motivated by the observation
that system \eqref{dint} is stationary
during the intertrial interval;
$x_1$ is either near 0 on rewarded trials
or some low value on error trials.
$x_2$ does not change on rewarded trials,
and saturates at some low value on error trials.
Figure~\ref{fig3} shows some activity during
the intertrial interval;
we model this additional activity
by a simple oscillation, $O = \sin(2 \pi f t + \phi)$,
where the frequency $f$ is low ($\le 0.4$Hz).
The final dynamics are
\begin{align} \label{dint-full}
  \dot{x}_1 &= u + E + O - R x_1 \nonumber \\
  \dot{x}_2 &= \alpha x_1 - R x_2.
\end{align}

Note that while there is experimental evidence for
a slow oscillations like $O$ in various brain regions
\citep{Buzsaki2004}, this injected oscillation
does not affect the performance of the network,
and is intended to model ACC activity
not being otherwise modelled in the current study.
This activity could reflect
other functions that the ACC is performing,
or could reflect the overall pace
of the simple RT task.

\section{Adaptive control circuit}

\citet{Narayanan2009} hypothesized
that the ACC activity analyzed in the present study
could be used to inhibit motor responses
until the time of the cue.
In order to verify that hypothesis,
we constructed a network that
performs the simple RT task
using the simple non-adaptive strategy
of responding to cues.
We then combined that network
with the double integrator network
such that the final model
is able to flexibly switch between
an aggressive timing strategy
and a conservative cue-response strategy
depending on the outcome of the previous trial.

\subsection{Cue-responding circuit}

The cue-responding circuit is composed
of three populations
of adaptive leaky integrate-and-fire neurons
(see Figure~\ref{fig4}).
The population labelled \textit{Trigger}
is sensitive to the cue.
The population labelled \textit{Holding}
tracks whether the lever is currently being held down.
The population labelled \textit{Press/Release}
causes lever presses and releases
through activation of different subsets
of its neurons.
The actual lever press and release
is delayed by $\sim$250ms,
the approximate amount of time
it takes for muscles to respond
to the activation of command neurons
in primary motor cortex.  % ???travis

@@

The context-driven design of this circuit
provides an alternative theory to
the prepotent inhibition model
proposed by \citet{Narayanan2009}.
They suggest that a motor command
is activated at the same time
as an inhibitory waiting signal,
and it is the removal of that inhibition
that produces the motor command.
This model proposes that
an excitatory context signal is activated,
and the pairing of that context signal
with the cue produces the motor command.
This leads to an alternative hypothesis
for the function of ACC activity:
the ACC tracks the behavioral context,
and is able to inhibit or excite motor cortex,
but does not always necessarily do so.

A weakness of the contextual cue-responding circuit
is that there is a fixed lower-bound
on the reaction time
based on the time taken to detect the cue
and for the delay between activation
of motor cortex and muscle activity.
Since the goal is to time the action
as close as possible to the time of the cue,
the cue-responding strategy may be too slow,
and so a predictive approach must be employed.

The double integrator model described
previously can implement such a strategy.
The state of the dynamical system being tracked by
the double integrator (equation \eqref{dint-full})
is approximately the same when the cue occurs.
By connecting the output of the double integrator
to the \textit{Trigger} population,
the double integrator can cause a lever release
when the state of the double integrator
is slightly before the predicted
time of the cue.

However, if the previous trial was an error,
the double integrator will not
be following the trajectory
that predicts the time of the cue,
and therefore will not cause a lever release.
The circuit then naturally
uses the conservative
cue-responding strategy.
Switching between these two strategies
depending on the outcome
of the previous trial
allows for adaptive control of action timing.

\section{Emulating aging effects}

The spiking neural implementation
of the double integrator model can be used
to emulate the effects of several pathologies.
As a proof of concept,
we emulate the effects of NMDA receptor degradation
in aging \citep{Magnusson1998}.

Feedforward connections
in the adaptive control model (Figure~\ref{fig4})
use a time constant of 10 ms,
which is consistent with the time constant
of the postsynaptic current
induced by uptake of neurotransmitter
by AMPA receptors \citep{Spruston1995}.
Recurrent connections use a time constant of 50 ms,
which is consistent with the PSC
of NMDA receptors \citep{Sah1990}.
To emulate the effects of aging on NMDA receptors,
we decrease the synaptic strengths
on these recurrent connections.

We hypothesize that this will have the same effect
as adding a small damping term, $D < 0$,
in equation \eqref{dint-full}, resulting in
\begin{align} \label{dint-damped}
  \dot{x}_1 &= u + E + O - R x_1 - D x_1 \nonumber \\
  \dot{x}_2 &= \alpha x_1 - R x_2 - D x_2.
\end{align}
This hypothesis can be tested directly
in the double integrator model,
and its effects on behavior tested
in the adaptive control circuit.

@@

The dynamical system described by equation \eqref{dint-full}
exhibits the desired behavior during simulation
of the simple RT task.
Figure~\ref{fig5} provides a breakdown of
the trajectory and vector field
throughout a correct, premature, and error trial.
Note that the oscillation is omitted
in this figure.
The final position of
the trajectory becomes the starting position
of the next trial, and therefore
provides information about
the outcome of the last trial.
It is clear from these trajectories
that the time of the cue
can be predicted,
as the ACC is only in the predictive state
when the cue occurs on correct trials.
It is also clear that the
outcome of the previous trials
is easily decoded based
on the state after the previous trial.
Note that premature trials
and late trials cannot be distinguished
on the next trial;
this is consistent with the results of
\citet{Narayanan2009}.

\subsection{The aged model loses adaptivity}

Figure~\ref{fig6} shows trajectories for
correct, premature, and late trials
for the aged model (i.e., the model with a damping term,
equation \eqref{dint-damped}).
As the damping term increases, the final position
of the trajectories on error trials
becomes closer to the final position of correct trials,
indicating that increased damping
makes distinguishing between correct and error trials
progressive more difficult.
Additionally, the distance (in state space) between
the state at which the cue occurs
and other states visited in the trajectory
decreases with increased damping,
which results in more difficulty
predicting the time of the cue
following correct trials as well.
Timing the foreperiod is therefore
a poor strategy with significant aging
(i.e., with $D \le 0.05$).

@@

The results indicate that the double integrator model
is allowing for adaptive control of action timing.
The ability to predict the time of the cue
produces faster reaction times at the cost
of additional premature errors.
The model that enables this adaptive control
has very similar neural dynamics
as the experimental ACC.

>>>>

\documentclass[11pt]{article}
\usepackage{amsmath,amsfonts,amsthm,amssymb}
\usepackage{setspace}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{extramarks}
\usepackage{chngpage}
\usepackage{caption}
\usepackage{soul}
\usepackage{namedplus}
\usepackage[usenames,dvipsnames]{color}
\usepackage{graphicx,float,wrapfig}
\usepackage{listings}
\usepackage{framed}
\usepackage{inconsolata}
\usepackage{hyperref}

@@

\pagestyle{plain}

@@

\setlength\fboxrule{1pt}
\linespread{1.5}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Make title
\title{A spiking neural integrator model of the adaptive control of
  action by the anterior cingulate cortex}
\date{Abbreviated title: A spiking model of ACC adaptive control}
\author{%
  Trevor Bekolay \\
  Centre for Theoretical Neuroscience \\
  University of Waterloo \\ ~ \\
  Mark Laubach \\
  The John B. Pierce Lab \\
  New Haven, CT \\ ~ \\
  Chris Eliasmith \\
  Centre for Theoretical Neuroscience \\
  University of Waterloo}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

@@

\textbf{Corresponding author:} Chris Eliasmith.
Email: \url{celiasmith@uwaterloo.ca}.
Centre for Theoretical Neuroscience, University of Waterloo.
200 University Avenue West, Waterloo, ON, N2L 3G1.

\textbf{Pages:} 24. \textbf{Figures:} 8. \textbf{Tables, multimedia, 3D models:} 0

\textbf{Word counts:} Abstract: 164, Introduction: 445, Discussion: 479

\textbf{Conflict of Interest}
The authors declare no competing financial interests.

\textbf{Acknowledgements}

Financial support: NSF 1121147 to ML;
NSERC Discovery, CRC, CFI, and OIT to CE;
and NSERC CGS-D to TB.
We thank Benjamine Liu for technical contributions
to the modeling effort and
Nandakumar Narayanan who recorded the neurons
that were analyzed by TB and ML.

\begin{abstract}
  Reaction times can be improved by preparing movements
  in advance and by learning the expected timing
  of action-imperative stimuli.
  Success or failure on the previous trial
  can be an important factor for determining
  whether an animal will attempt to time
  the stimulus or simply wait for it to occur
  before initiating action.
  The anterior cingulate cortex (ACC) has been implicated
  in enabling the top-down control of action
  depending on the outcome of the previous trial.
  Analysis of ACC activity suggests that
  neural integration is a key mechanism
  for adaptive control in precisely timed tasks.
  We show through simulation
  that a spiking neural circuit
  consisting of coupled neural integrators
  captures the neural dynamics
  of the experimentally recorded ACC.
  We expand on this coupled integrator model
  to construct a spiking neural circuit that
  performs a reaction time task by following
  either a timing or cue-response strategy,
  and show that it performs a simple reaction time task
  with similar reaction times as experimental subjects.
\end{abstract}

\linespread{1.5}

@@

The anterior cingulate cortex (ACC) has been implicated
in a wide variety of functions,
including conflict monitoring
\cite{Sheth2012,Aarts2009,Botvinick2004,VanVeen2004},
performance monitoring
\cite{Brown2005,Carter1998,Fecteau2003,Modirrousta2008,Rushworth2004,Histed2009,Hyman2013,Horst2012},
anticipatory control \cite{Aarts2008,Koyama2001},
arousal of the sympathetic nervous system
\cite{Critchley2003,Luu2003},
and control of action timing
\cite{Muir1996,Mulert2003,Naito2013,Narayanan2009,Risterucci2003}.
Conceptual models have been proposed
that would allow the ACC to be involved
in several of these functions,
depending on behavioral context
(e.g., \citenoparens{Botvinick2004,Luu2003}).
However, while there have been
attempts at modeling the ACC
using artificial neural networks
\cite{Botvinick2004,Brown2005},
there are currently no mathematical
accounts of ACC function that would
explain the myriad of behaviors
modulated by the ACC.
As well, no existing models
have been implemented with spiking neurons,
and therefore no models can be analyzed
with the same methods used
to analyze single neuron recordings.

In this paper, we propose that a
two-dimensional dynamical system
can account for ACC activity
related to error monitoring and motor action timing.
We test the model in a simple reaction time task
with a predictable foreperiod and
show that the mathematical model
enables precise timing of actions,
and changes behavior depending
on the outcome of the last trial.
The dynamical system is based
on a multi-dimensional attractor
network proposed by \citetext{Singh2006},
and is implemented in a spiking neural network
using the Neural Engineering Framework
(NEF; \citenoparens{Eliasmith2003}).
We show, using identical analyses
to those employed on the experimental data,
that simulated neural activity from this circuit
closely matches recorded ACC activity
and that the model can be combined with
a simple motor control model
to produce behavior consistent
with behavioral studies of simple reaction time tasks.

Our model also provides an alternative hypothesis
of how the ACC accomplishes the control of action timing.
\citetext{Narayanan2009} hypothesized
that ACC activity inhibits motor responses
until the time of the cue.
Our model instead proposes that
ACC activity encodes the duration of the action
and that inhibition is carried out elsewhere.

In this paper we apply our
mathematical account of ACC function
to explain error monitoring and
the control of action timing.
However, we suggest that the dynamical modelling
method we employ is general and can be extended
to explain other theories
of high-level ACC function.

\subsection{The reaction time task} \label{sec:simple-rt}

Twelve male Long-Evans rats
were trained to press down on a lever
for a fixed amount of time (the ``foreperiod'';
1 second for this study)
using standard methods \cite{Laubach2000}.
After the foreperiod, a cue is presented
(an auditory tone).
The subject then has 0.6 seconds (the response window)
in which to release the lever.
If the lever is released in that time,
the trial is a success, and the subject is rewarded
with water.
If the lever is released during the foreperiod,
the trial is classified as a ``premature error,''
and is penalized by a time-out with the house lights extinguished.
If the lever is not released within the response window,
the trail is classified as a ``late error,''
and is also penalized by a time-out with the house lights extinguished.
Figure~\ref{fig1} presents a schematic view of the task.
Subjects were considered to have learned the task
when they performed in correctly $> 60 \%$
of the time.

The same reaction time task was simulated
in the Nengo simulation environment
using a finite state machine to track
task state, and scalar values between -1 and 1
to indicate the state of house lights,
lever position, auditory cue, and reward.
The Nengo scripts simulating this environment
are available at
\url{https://github.com/tbekolay/jneurosci2013}.

\subsection{Data analysis}

\subsubsection{Neural data}

The experimental data analyzed in this study is the same
as that in \citetext{Narayanan2009}, in which
the authors recorded the anterior cingulate cortex (ACC) and motor cortex
of rodents performing this simple reaction time task.
(The ACC recordings were made in the prelimbic cortex,
which is a component of the ACC.
The area was referred to as the dorsomedial PFC
in the original paper.)
\citeauthor{Narayanan2009} used principal component analysis
(PCA) to characterize common firing patterns
in the ACC during the task.

In the PCA, only neurons with an average
firing rate over 1 Hz are considered.
Spike trains from 4 seconds prior to and 4 seconds after
each press event on a correct trial are isolated.
Each perievent spike train is binned in 1 ms bins and is then
convolved with a Gaussian filter ($\sigma = $ 25 ms;
PCs are consistent with many other $\sigma$ values).
These spike density functions
are normalized to Z-scores,
and then averaged over all trials to
produce a matrix in which each row is
the normalized average response of a neuron
over the perievent epoch.
Singular value decomposition
is performed on that matrix
to isolate the principal components,
which are then normalized to Z-scores.
The amount of variance accounted for
by each principal component
is computed as the square of the component's eigenvalue
over the sum of all squared eigenvalues
($s_i^2 / \sum s^2$).

As shown in Figure~\ref{fig2},
the first two principal components
of the experimental data
are modulated around the task events
and appear to track
whether the subject has pressed the lever and
the relative amount of time
that the lever has been pressed.
\citeauthor{Narayanan2009} also found that the neural activity
in the ACC changed significantly depending on
the outcome of the previous trial;
the first two principal components
change significantly after errors,
and appear to contain information that an error occurred.
This information would be necessary
in order to adapt behavior based on the last trial.

Interestingly, in the post-correct case,
the first principal component
is highly correlated with the integral
of the second principal component,
and vice versa.
In the post-error case,
these correlations are significantly weaker.
This result suggests that
the ACC is performing some kind of integration,
and that integration depends
on prior behavioral outcomes.

\subsubsection{Behavioral data}

In both the experimental and theoretical studies,
the following relevant event times are recorded:
lever press, lever release, reward delivery,
trigger stimulus onset, and house light extinguishing.
Sequences of event times were used to
identify the result of each trial.
Filtering the set of event timestamps
for these sequences identifies each trial outcome.
Reaction times are calculated
only for correct trials,
and are defined as the difference between
trigger stimulus onset and lever release.

\subsection{Double integrator model}

The double integrator model
was proposed in the context of working memory
by \citetext{Singh2006}.
Their model uses multidimensional neurons
in integrative populations
to exhibit time-varying activity
during a vibrotactile discrimination task.
It relies on two coupled integrators,
whose activity resembles
the two strong principal components
in the ACC data.
The model employs
the Neural Engineering Framework (NEF; \citenoparens{Eliasmith2003})
to implement a spiking neural network
that captures the dynamics of the
integrative dynamical system.
We hypothesize that a model
similar to that proposed by \citeauthor{Singh2006}
will exhibit the same neural dynamics
as the experimental ACC in the simple RT task.

\subsubsection{Dynamics} \label{sec:dynamics}

The dynamics of the double integrator model
are captured in the following dynamics equations:
\begin{align*}
  \dot{x}_1 &= u \\
  \dot{x}_2 &= \alpha x_1,
\end{align*}
where $u$ represents an input signal
and $\alpha$ is the strength of
the connection between the two integrators.
This results in a two-dimensional system
in which $x_1$ integrates its input,
maintaining the accumulated value over time, and
$x_2$ integrates the value of $x_1$.
\citetext{Singh2006} have proposed that this system
enables time tracking.
Because $x_2$ integrates $x_1$,
the amount of time elapsed since
$x_1$ changes can be determined
based on how much $x_2$ has changed.

In the simple RT task,
we want to keep track of how much time
has elapsed since the subject
has pressed the lever.
While the lever press event could be
used as the input to the system,
since we aim to embed this system
in a spiking neural model,
we instead use the activity of the motor neurons
that effect the lever press as
the input, $u$.
The system is also modified by signals
representing the outcome of a particular trial;
on a correct trial, reward delivery ($R > 0$)
resets the system to a starting state.
On an error trail, a second input signal ($E < 0$)
drives $x_1$ to a low state.
When implemented in a spiking network,
this causes $x_2$ to saturate at a low value,
effectively breaking the integration,
as is seen in Figure~\ref{fig2} in post-error trails.
This results in the following overall dynamics.
\begin{align} \label{dint}
  \dot{x}_1 &= u + E - R x_1 \nonumber \\
  \dot{x}_2 &= \alpha x_1 - R x_2.
\end{align}
See Figure~\ref{fig3} for a graphical
depiction of the double integrator model.

The dynamics of Equation \eqref{dint}
are predictable for certain values
of $\alpha$, $E$, and $R$.
$\alpha$ defines how quickly
the system drifts during the task;
we set it such that $x_2 \approx 1$
at the time of release,
which occurs when $\alpha = 0.04$.
If $R$ is sufficiently big,
then when reward is delivered
the system will be driven to the origin.
This occurs when $\int_0^{t_R} R\,dt \ge 1$.
We set the length of reward delivery,
$t_R$, to 1 second to reflect the pace
of the experimental subjects,
and we set the reward magnitude,
$R$, to 0.15 so that $\int_0^{t_R} R\,dt \ge 1$.
Larger values of $R$ would be functionally identical,
as $x_1 = x_2 = 0$ when $R \ge 0.15$ for $t_R$ seconds.
If $E$ is sufficiently low,
then when error is detected,
the system will be driven to
the lowest $x_1$ value that can
be represented by the population
of spiking neurons.
This exploits the saturating nature
of spiking neurons in order to
obtain stability in this
mathematically unstable system.
If $\int_0^{t_E} E\,dt \le -1$,
then the system will saturate
at a low value for $x_1$.
We set $t_E = 2 s$ and $E = -0.06$.

To summarize, there are five parameters
in this dynamical system:
$t_R$ and $t_E$ are set based on the pace
of experimental subjects;
$R$ and $E$ have little behavioral
consequence if $R \ge 0.15$ and
$E \le -0.06$ and are set
to match the experimental principal components;
$\alpha$ is a free parameter
that critically affects the model's success
and is set such that $x_2 \approx 1$
at the time of release.

In addition to the above system,
we inject a slowly oscillating signal into $x_1$.
Having made the parameter choices
described above,
the system is stable during the intertrial
interval in both post-correct and post-error
cases.
However, the experimental data shown in
Figure~\ref{fig2} exhibits some activity during
the intertrial interval;
we capture this additional activity
by introducing a simple oscillation,
$O = \sin(2 \pi f t + \phi)$,
where the frequency $f$ is low ($\le 0.4$Hz).
While there is experimental evidence for
a slow oscillations like $O$ in various brain regions
\cite{Buzsaki2004}, this injected oscillation
does not affect the behavioral performance of the network,
but instead models ACC activity
not being explicitly modelled in
the remainder of the circuit.
This activity could reflect
other functions that the ACC is performing,
or could reflect the overall pace
of the simple RT task.
The final dynamics are
\begin{align} \label{dint-full}
  \dot{x}_1 &= u + E + O - R x_1 \nonumber \\
  \dot{x}_2 &= \alpha x_1 - R x_2.
\end{align}

Equation~\eqref{dint-full} is implemented
in a spiking neural network using
the principles of the Neural Engineering Framework
(NEF; \citenoparens{Eliasmith2003}).
The NEF enables the simulation of dynamical systems
like Equation~\eqref{dint-full} by
making the assumption that populations of spiking neurons
represent real-valued vectors,
and using a least-squares optimal method
to transform those representations
through the connections between
populations of neurons.
A population can have recurrent connections
in order to produce dynamics,
such as the neural integrators
employed in this study.

The NEF's representation scheme
is similar to population coding,
as proposed by \citetext{Georgopoulos1986},
but extended to $n$-dimensional vector spaces.
Each neuron in a population is sensitive
to a particular direction,
called the neuron's \textit{encoder}.
The activity of a neuron can be expressed as
\begin{equation}
  a = G[\alpha \mathbf{e} \cdot \mathbf{x} + J_{bias}],
\end{equation}
where $G[\cdot]$ is a nonlinear neural activation function,
$\alpha$ is a scaling factor (gain) associated with the neuron,
$\mathbf{e}$ is the neuron's encoder,
$\mathbf{x}$ is the vector to be encoded, and
$J_{bias}$ is the background current of the cell
when $\mathbf{x} = 0$.
The currently encoded value, $\mathbf{\hat{x}}$,
can be estimated linearly:
\begin{equation}
  \mathbf{\hat{x}}(t) = \sum_i \mathbf{d}_i a_i(t),
\end{equation}
where $\mathbf{d}_i$ is the neuron's decoder,
and $a_i$ is activity of neuron $i$.

Decoders are found through a least-squares minimization
of the difference between the decoded estimate
and the actual encoded vector.
\begin{equation}
  \mathbf{d} = \Upsilon^{-1} \Gamma \hspace{1.8em}
  \Gamma_{ij} = \int a_i a_j dx \hspace{1.8em}
  \Upsilon_j = \int a_j \mathbf{x} dx.
\end{equation}

Connection weights to linearly transform these encoded vectors
can then be computed as
\begin{equation}
  \omega_{ij} = \alpha_j \mathbf{e}_j L \mathbf{d}_i,
\end{equation}
where $i$ indexes the input population,
$j$ indexes the output population,
and $L$ is a linear operator.

% We only do linear transformations I think?

% Connection weights computing a function
% between two populations connected in a feedforward manner
% $f(\mathbf{x})$ can be determined by
% solving for a set of decoding weights
% for that function,
% \begin{equation}
%   \mathbf{d}^{f(\mathbf{x})} = \Upsilon^{-1} \Gamma \hspace{1.8em}
%   \Gamma_{ij} = \int a_i a_j dx \hspace{1.8em}
%   \Upsilon_j = \int a_j f(\mathbf{x}) dx,
% \end{equation}
% and then computing the following weight matrix, $\omega$;
% \begin{equation}
%   \omega_{ij} = \alpha_j \mathbf{e}_j L \mathbf{d}^{f(\mathbf{x})}_i,
% \end{equation}
% where $i$ indexes the input population,
% $j$ indexes the output population,
% and $L$ is a linear operator.

Dynamical systems of the form
$\dot{\mathbf{x}} = A(\mathbf{x}) + B(\mathbf{u})$
can be created by connecting a population
to itself with the linear operator
$L = A$ on the recurrent connection,
and receiving input $\mathbf{u}$
from other populations and
setting $L = B$ on those connections.

These principles can be used for any neuron model,
$G[\cdot]$. In this study, we use the
adaptive leaky integrate-and-fire (ALIF) model
to match \citetext{Singh2006}.
For each instance of each model,
the parameters of each ALIF neuron
is randomly selected from within
a biologically constrained range.
The encoders, $e_i$, are randomly selected
directions in the encoded vector space.
All other parameters in the simulation
are computed from these parameters,
or are described in section \ref{sec:dynamics}.

\citetext{Narayanan2009} hypothesized
that ACC activity could be used
to inhibit motor responses
until the time of the cue.
An alternative view is that neuronal activity
serves to encode the duration of the action
and that inhibition is carried out elsewhere.
To examine these hypotheses,
we constructed a network that
performs the simple RT task
using the non-adaptive strategy
of responding to cues.
We then combined that network
with the double integrator network
to give an overall ``adaptive control circuit''
that is able to switch between
an aggressive timing strategy
and a conservative cue-response strategy
depending on the outcome of the previous trial.

\subsubsection{Cue-responding circuit} \label{sec:cue-responding}

The cue-responding circuit is composed
of three populations
of adaptive leaky integrate-and-fire neurons
(see Figure~\ref{fig3}, grey box only).
The population labelled \textit{Trigger}
is sensitive to the cue.
The population labelled \textit{Holding}
tracks whether the lever is currently being held down.
The population labelled \textit{Press/Release}
causes lever presses and releases
through activation of different subsets
of its neurons.
The actual lever press and release
is delayed by $\sim$200ms,
the approximate amount of time
it takes for muscles to respond
to sensory input \cite{Kawato1999}.

@@

The context-driven design of this circuit
provides an alternative theory to
the prepotent inhibition model
proposed by \citetext{Narayanan2009}.
They suggest that a motor command
is activated at the same time
as an inhibitory waiting signal,
and it is the removal of that inhibition
that produces the motor command.
This model proposes that
an excitatory context signal is activated,
and the pairing of that context signal
with the cue produces the motor command.
This leads to an alternative hypothesis
for the function of ACC activity:
the ACC tracks the behavioral context,
and is able to inhibit or excite motor cortex,
but does not always necessarily do so.

\subsubsection{Adaptive control circuit}

A weakness of the contextual cue-responding circuit
is that there is a fixed lower-bound
on the reaction time
based on the time taken to detect the cue
and for the delay between activation
of motor cortex and muscle activity.
Since the goal is to produce the action
as close as possible to the time of the cue,
the cue-responding strategy may be too slow,
and so a predictive approach must be employed.

The double integrator model described
previously can be combined with the
cue-responding circuit to implement such a strategy
(see Figure~\ref{fig3}).
The state of the dynamical system being tracked by
the double integrator (Equation [\ref{dint-full}])
is approximately the same each trial when the cue occurs.
By connecting the output of the double integrator
to the \textit{Trigger} population,
the double integrator can cause a lever release
when the state of the double integrator
is slightly before the predicted
time of the cue.

However, if the previous trial was an error,
the double integrator will not
be following the trajectory
that predicts the time of the cue,
and therefore will not cause a lever release.
The circuit then naturally
uses the conservative
cue-responding strategy.
Switching between these two strategies
depending on the outcome
of the previous trial
allows for adaptive control of action timing.
A possible extension to this model
would incorporate synaptic plasticity
to change the recurrent connections such that
the cue prediction would be better
on subsequent trials.

@@

Here we describe the results of solving
the dynamical system in equation \eqref{dint} directly
(i.e., not using neurons).
This provides a general characterization
of the expected low-dimensional dynamics
of the high-dimensional neural simulation.
Figure~\ref{fig4} demonstrates that
this dynamical system exhibits
the behavior described in section \ref{sec:dynamics} during
correct, premature, and late trials.
Note that this figure depicts equation \eqref{dint},
which does not have the oscillatory term $O$.
The final position of
the trajectory becomes the starting position
of the next trial, and therefore
provides information about
the outcome of the last trial.

It is clear from these trajectories
that the time of the cue
can be predicted,
as the ACC is only in the predictive state
when the cue occurs on correct trials.
It is also clear that the
outcome of the previous trials
is easily decoded based
on the state after the previous trial.
Note that premature trials
and late trials cannot be distinguished
on the next trial;
this is consistent with the results of
\citetext{Narayanan2009}.

These theoretical results suggest that
the underlying, low-dimensional dynamics
of the postulated circuit
should be appropriate
for capturing the qualitative features of ACC activity.
We now examine
both the qualitative and quantitative fit
of the high-dimensional spiking network
to the recorded empirical data.

@@

Our results indicate that the double integrator model
allows for adaptive control of action timing.
The ability to predict the time of the cue
produces faster reaction times at the cost
of additional premature errors.
The model that enables this adaptive control
has very similar neural dynamics
as the experimental ACC.
